Ricci curvature of the rooted phylogenetic subtree-prune-regraft graph

### 2015-02-09 curvature.tex
close relationship between lazy and uniform curvature

\begin{proof}
	$T$ and $S$ differ by one SPR move that moves a subtree $R$.
	Let $U$ be a tree in $N(T)$ and $N(S)$.
	Then either (1) $T$ and $U$ differ in the location of $R$, or (2) $T$ and $U$ differ in the location of another subtree $P$.
	In the latter case, the facts that $T$ and $S$ differ only in the location of $R$ and $\dspr{T,P} = \dspr{S,P} = 1$, imply that $T|(X \setminus L(P) - S|(X \setminus L(P)$.
	Then leaves $r' \in R$, $p' \in P$, and $q' \in Q$, for some subtree $Q$, form a triple of $T$ and a different triple in $S$.
	This incompatible triple can be resolved in at most $6n - 17$ ways, the maximum of which is reached when $P$, $Q$, and $R$ are themselves a ``triple'' of subtrees.
	By Lemma~\ref{ }\todo{need unique move lemma}, each of the subtrees is assigned to at most $2n-6$ unique moves.
	Moreover, one additional overlapping move also moves one of the subtrees (that of the aunt of the LCA of the three subtrees).
	The number of shared neighbors is thus at most $3(2n-6) + 1 = 6n-17$.
	Note that this bound is tight when, for example, $T$ and $S$ are ladders with a different configuration of 3 leaves at maximum depth.
\end{proof}

@@

\begin{lemma}
	The maximum curvature of the uniform random walk between two adjacent trees is $\frac{6n-17}{3n^2-13n+14}$. \todo{lazy random walk is $\frac{6n-15}{3n^2 - 13n + 14}$}
\end{lemma}
\begin{proof}
Let $T$ and $S$ be two adjacent trees.
The maximum curvature occurs when their neighborhoods have maximum overlap, are of equal size, and all other tree pairs are at distance 1.
By Lemma~\ref{lem:shared_neighbors} the maximum overlap is $6n-17$.
The maximum curvature is thus $1 - \frac{\degree{T} - 6n-17}{\degree{T}} = \frac{6n - 17}{\degree{T}}$. This is maximized when $\degree{T}$ is as small as possible, that is $T,S$ are ladders and $\degree{T} = 3n^2 - 13n + 14$.
\end{proof}

>>>>

\begin{proof}
	$T$ and $S$ differ by one SPR move that moves a subtree $R$.
	Let $U$ be a tree in $N(T)$ and $N(S)$.
	Then either (1) $T$ and $U$ differ in the location of $R$, or (2) $T$ and $U$ differ in the location of another subtree $P$.
	In the latter case, the facts that $T$ and $S$ differ only in the location of $R$ and $\dspr{T,P} = \dspr{S,P} = 1$, imply that $T|(X \setminus L(P)) = S|(X \setminus L(P))$.
	Then leaves $r' \in R$, $p' \in P$, and $q' \in Q$, for some subtree $Q$, form a triple of $T$ and a different triple in $S$.
	This incompatible triple can be resolved in at most $6n - 17$ ways, the maximum of which is reached when $P$, $Q$, and $R$ are themselves a ``triple'' of subtrees.
	By Lemma~\ref{ }\todo{need unique move lemma}, each of the subtrees is assigned to at most $2n-6$ unique moves.
	Moreover, one additional overlapping move also moves one of the subtrees (that of the aunt of the LCA of the three subtrees).
	The number of shared neighbors is thus at most $3(2n-6) + 1 = 6n-17$.
	Note that this bound is tight when, for example, $T$ and $S$ are ladders with a different configuration of 3 leaves at maximum depth.
\end{proof}

@@

\begin{lemma}
	The maximum curvature of the uniform random walk between two adjacent trees with $n$ leaves is $\frac{6n-17}{3n^2-13n+14}$. \todo{lazy random walk is $\frac{6n-15}{3n^2 - 13n + 14}$}
\end{lemma}
\begin{proof}
Let $T$ and $S$ be two adjacent trees.
The maximum curvature occurs when their neighborhoods have maximum overlap, are of equal size, and all other tree pairs are at distance 1.
By Lemma~\ref{lem:shared_neighbors} the maximum overlap is $6n-17$.
The maximum curvature is thus $1 - \frac{\degree{T} - 6n-17}{\degree{T}} = \frac{6n - 17}{\degree{T}}$. This is maximized when $\degree{T}$ is as small as possible, that is $T,S$ are ladders and $\degree{T} = 3n^2 - 13n + 14$.
\end{proof}

We next bound the effect of laziness on curvature. Let $\kappa_{l,p}$ be the curvature of a lazy random walk which moves with probability $p$ and remains stationary with probability $1 - p$. 
For example, $\kappa_{l,\frac{1}{4}}$ moves with probability $\frac{1}{4}$ and remains stationary with probability $\frac{3}{4}$. \todo{should be defined earlier, particularly the part where the curvature is divided by $p$.}

\begin{lemma}
	Let $T$ and $S$ be rooted phylogenetic trees with $n$ leaves.
	Then:
	\begin{enumerate}
		\item	$\kappa_{l,p}(T,S) = \kappa_{l,0}(T,S)$, if $\dspr{T,S} > 1$, and
		\item	$\kappa_{l,0}(T,S) \le \kappa_{l,p}(T,S) \le \kappa_{l,0}(T,S) + \frac{2}{\max(\degree{T}, \degree{S})}$, if $\dspr{T,S} = 1$.
	\end{enumerate}
\end{lemma}

\begin{proof}
	We first prove the lower bound, that is $\kappa_{l,0}(T,S) \le \kappa_{l,p}(T,s)$.
	Let $W_1(T,S)$ be the mass transport cost in the uniform case, and $W_1'(T,S)$ be the same for the lazy case with parameter $p$.
	Recall that $\kappa_{l,0}(T,S) = 1 - \frac{W_1(T,S)}{\dspr{T,S}}$, and $\kappa_{l,p}(T,S) = (1 - \frac{W_1'(T,S)}{\dspr{T,S}}) / p$.
	Observe that $$W_1'(T,S) \le pW_1(T,S) + (1-p)\dspr{T,S},$$ by taking an optimal uniform mass assignment spread over $p$ of the mass and moving the remainder from $T$ to $S$ (or vice versa).
	Then:
	\begin{align}
		\kappa_{l,p}(T,S) &= (1 - \frac{W_1'(T,S)}{\dspr{T,S}}) / p \\
		&\ge (1 - \frac{pW_1(T,S) + (1-p)\dspr{T,S}}{\dspr{T,S}}) / p \\
		&\ge (\frac{1}{p} - \frac{W_1(T,S)}{\dspr{T,S}} - \frac{1-p}{p}) \\
		&\ge 1 - \frac{W_1(T,S)}{\dspr{T,S}} \\
		&\ge \kappa_{l,0}(T,S)
	\end{align}

	For the upper bound, we observe that $$W_1'(T,S) \ge p(W_1(T,S) + (1-p)\left(\dspr{T,S}\right) - \frac{2}{\max(\degree{T}, \degree{S})},$$ as at most $\frac{1}{\max(\degree{T},\degree{S})}$ of the mass can remain at each of $T$ and $S$, paired with the lazy remainder.
	The upper bound then follows analogously to the lower bound.
	Moreover, no mass can remain at $T$ or $S$ when $\dspr{T,S} > 1$, in which case the curvatures are equal.
\end{proof}
